Financialisation of Commodity Markets \\\large{Co-movement Behind-the-Scenes}

### 2024-10-23 docs/paper.tex
Editorial polishing

\begin{abstract}
In the early 2000s institutional investors entered the commodity futures
markets en masse with passive, long only, index type positions in sharp
contrast with those typically assumed by traditional expert
participants. A heated public debate soon erupted over the perceived
consequences of the phenomenon---commonly referred to as
``financialisation''---and, in response to immediate policy concerns,
the matter was thrust onto the academic sphere as a burning issue. With
the benefit of hindsight it now seems that the academic debate was
framed rather narrowly, used contentious research methods, and
eventually led to regulatory changes that were therefore perhaps
unwarranted. In contrast, we take a broader approach where we consider a
large cross-section of liquid commodities as suggested by the nature of
financialisation that comprehends commodity futures as an asset class.
Using a bespoke asset pricing factors based framework we study the issue
through the lens of co-movement and find that the phenomenon had
ontological consequences for the commodity complex with its impact
extending beyond the mechanical effects induced by indexation. The onset
of the financial crisis and the monetary policy regimes that followed,
on the other hand, seem to have set off a motion of reversion to legacy
pre-financialisation fundamentals.
\end{abstract}

@@

The last twenty years have seen major upheavals in the commodity markets
with the onset of financialisation\footnote{Starting around 2004, with a
  view of commodity markets as an emerging asset class and in an effort
  to hedge against inflation and seek diversification benefits
  \citep{buyuksahin_speculators_2014, singleton_investor_2013},
  institutional investors sent forth an unprecedented flow of capital
  into commodity futures markets. This phenomenon is commonly referred
  to as the financialisation of commodities
  \citep{domanski_financial_2007}.} as well as the period of the 2008
financial crisis and its aftermath. Financialisation appears to have
affected multiple economic sectors, including agriculture and energy,
and its impact has been hotly debated in the policy, legislative,
regulatory, and academic spheres. A rigorous legislative scrutiny
accompanied by a vigorous policy and academic debate across a number of
disciplines about the perceived effects on commodity prices of the
unprecedented inflow of institutional funds brought into commodity
futures markets by new index type investors\footnote{See section
  Section~\ref{sec-background} for a detailed discussion.} eventually
led to regulatory changes in several individual markets.\\
In response to immediate policy concerns, the academic debate was
initially framed rather narrowly around adequacy of speculation---the
burning issue of the day---at the individual commodity market level and
the ensuing academic analysis focused on the more mechanical effects of
financialisation. With the benefit of hindsight, this approach seems to
lack a fundamental understanding of the phenomenon it endeavoured to
study and which consequences it attempted to address.\\
Financialisation was a phenomenon global in nature that, through the
channel of index investment, affected the whole cross-section of liquid
commodities at once, eventually spurring its transition away from a
collection of idiosyncratic markets to its emergence as an asset class.
Focusing on individual markets in this context thus seems limited. Most
of the studies underpinning the regulatory response to financialisation
took the form of pairwise correlation, causation and to some extent
co-integration analysis, all of which are idiosyncratic in nature for
they study single pairs of individual assets. Besides, some of these
techniques since proved controversial and seem to produce results the
quality of which may fall short of the standards that may reasonably be
expected to lay the ground for legislative action. Although appealing,
for it comes with the promise of prompt ``significant'' results, this
approach misses a critical corollary of the substantial advent of a new
category of players in a (set of) market(s): the alteration of
fundamentals it potentially induces, best observed in this context
through the lens of commercial hedging pressure
\citep{keynes_treatise_1930, hicks_value_1939}.\\
While essential to the understanding of the phenomenon, the global
nature of financialisation and its impact on market fundamentals
therefore appear to have been originally overlooked and another
approach, perhaps more refined, seems warranted.

@@

These entered the commodity spectrum en masse with across-the-board
index type positions propping the transition of the complex from a
collection of idiosyncratic markets towards a more congruent ensemble
akin to an asset class. In this context of heightened cross-asset market
participation the issue of interconnectedness stands out as paramount
and we strive to investigate it by first examining the correlation
structure of our sample set of assets at the global, country, sector and
subsector levels. Our results indicate higher interdependence within the
full cross-section of commodities over the financialisation period,
consistent with both the empirical and theoretical literature. The
increase appears to be driven by the US metals sector, a new observation
to the best of our knowledge, and, unlike individual commodity returns,
does not seem to be regime-driven, as a similar pattern of results is
observed across phases of \textbf{backwardation} and \textbf{contango}.

@@

The returns pull-back continues into the post-crisis period,
interestingly amidst a lower volatility environment, suggesting that
investment outflows from the commodity futures complex have had a lower
impact on volatility than the corresponding inflows witnessed during
financialisation. The pattern of reversion toward the
pre-financialisation Keynesian paradigm continues stronger with mean
returns higher during phases of \textbf{backwardation} relative to
\textbf{contango} and the difference more pronounced than over the
crisis period. In contrast to the crisis period when they seemed to be
converging, the pricing performance of CHP and \textbf{CHP} now part
way, seemingly reverting to the Keynesian paradigm observed
pre-financialisation, further suggesting a reversion to legacy commodity
specific dynamics. The pull back extends to the correlation structure
with an across-the-board decrease in average pairwise correlations down
to levels below those observed during financialisation. The average
market factor beta for the entire cross-section shows a similar pattern,
dropping down below the level observed during financialization with the
decline seemingly driven by the metals sector, particularly the GB
traded segment. The pattern extends to the pricing performance of all
factors that drops below levels observed during financialisation while
the gap between the legs of the long-short factors remains narrower than
observed during that period.\\
Moreover, both the average correlations and the pricing performance of
the market factor are considerably stronger during \textbf{contango}
phases, unlike previous periods, which showed no substantial difference
between regimes with especially pronounced for metal assets globally,
both US and GB traded. These results document the emerging
role/dominance of the \textbf{contango} regime post-crisis and could
provide a valuable basis for further research.

@@

A world-wide debate ensued about the role of index funds in commodity
markets. The first responses to the 2007/2008 crisis of escalating food
and energy prices took the form of policy reports, many of which
reasoned that the growth of commodity index funds came along with an
influx of largely speculative capital that was responsible for driving
commodity prices beyond their historic highs
\citep{deschutter_food_2010, gilbert_speculative_2010, herman_not_2011, schumann_hunger_2011, unctad_global_2009}.\\
Early US Senate investigations on the matter drew on pricing and trading
data supplied by the CFTC as well as interviews with numerous experts
including hedge fund manager Michael Masters who linked the growing
presence of index investors in US agricultural and energy markets to the
observed surges in both futures and cash prices
\citep{masters_testimony_2008, masters_accidental_2008}. This contention
has since come to be known as the Masters' hypothesis and was largely
endorsed by the final US Senate report on the issue
\citep{senate_excessive_2009}.\\
At this stage, the question thrust on the academic sphere was whether
``excessive speculation'' (understood as the market activity peculiar to
index type investors) was linked to escalating energy and agricultural
commodity prices. A number of ensuing academic studies, most of which
used commodity specific correlation and causality based analysis have
since disproved this contention: \citet{irwin_devil_2009},
\citet{sanders_adequacy_2010}, \citet{sanders_impact_2011},
\citet{sanders_new_2011}, \citet{irwin_commodity_2013},
\citet{brunetti_commodity_2014}, \citet{hamilton_effects_2015} and
\citet{bruno_financialisation_2017} investigated the issue in the
context of the agricultural markets; meanwhile,
\citet{buyuksahin_speculators_2011}, \citet{tokic_speculation_2012},
\citet{fattouh_role_2013}, \citet{kilian_role_2014},
\citet{knittel_simple_2016} and \citet{manera_modelling_2016} examined
the energy markets while \citet{bohl_does_2013}, \citet{kim_does_2015}
and \citet{boyd_prevalence_2016} studied both energy and agricultural
markets and \citet{irwin_index_2011},
\citet{irwin_financialisation_2012} as well as
\citet{stoll_commodity_2011} examined the commodity markets in general.
Other studies have underlined various alterations in commodity pricing
dynamics over the period. \citet{gilbert_role_2014} reject the view that
financialisation has not had any effects in the grains markets and
demonstrate that trades originated by financial market participants, and
specifically index investors, can move prices but tend to be typically
volatility-reducing. \citet{juvenal_speculation_2015} contend that the
oil price increase between 2004 and 2008 was mainly driven by the
strength of global demand but that the financialisation process of
commodity markets also played a role. Likewise,
\citet{henderson_new_2015} show that non-information-based financial
investments have important impacts on commodity prices.
\citet{cheng_financialisation_2014}, on the other hand, show that in the
case of index commodities, financialisation has transformed the risk
sharing, and information discovery functions of commodity futures
markets.

@@

For each period, the return series for the individual US-traded assets
are regressed on both the relative change on their respective CHP series
and that on the cross-sectional CHP independently. The resulting
\(R^{2}\) values are then averaged by sector and subsector.
Additionally, pairwise correlations are computed for each combination of
assets across the periods of interest and for each individual year
within the full sample period. These correlation coefficients are
subsequently averaged across the entire cross-section of assets, as well
as by country, sector, and subsector. For each period, the coefficients
are also calculated and averaged across phases of low and high
cross-sectional CHP.

@@

(XLME)---show a lower figure with five significant figures---Crude

@@

The onset of financialisation thus seems to have engendered a change in
the nature of hedger's behaviour, at the aggregate level in particular,
taking it away from the traditional Keynesian view of risk shifting.
This phenomenon has been noted in earlier theoretical and empirical work
\citep{danthine_information_1978, stout_why_1998} in different contexts
from which two competing models of behaviour have emerged.\\
The information arbitrage model
\citep{grossman_impossibility_1980, danthine_information_1978, kyle_informed_1989}
implies that hedgers may often be seeking out counterparties to trade
with rather than being purely passive. This issue is also raised in both
\citet{cheng_financialisation_2014} and \citet{stulz_rethink_1996} who
point out that hedgers may be taking a view on prices just as
speculators do. In fact, by hedging away some of their risk, hedgers are
able to speculate more heavily based on their disagreement against
speculators regarding futures price movement
\citep{simsek_speculation_2013}, which fits into a heterogeneous
expectation theory of speculation\footnote{\citet{hirshleifer_speculation_1975};
  \citet{hirshleifer_reply_1976}; \citet{hirshleifer_theory_1977}}. As
\citet{stout_why_1998} points out, this disagreement-based theory of
trading is one of the main reason the public and the law disapproves
speculation as this form of trading is regarded as
non-productive\footnote{\citet{duffie_challenges_2014} also discusses
  some of the challenges faced by a policy treatment of speculative
  trading motivated by differences in beliefs.}.\\
In the second model hedgers are seen as liquidity providers as in
\citet{kang_tale_2020} with this explanation also consistent with the
nature of financialisation which led to the arrival of long term, long
only investors who require substantial liquidity to roll-over their
positions. Table~\ref{tbl-time-reg-US-CHP} supports this view. We
regress commodity futures front returns on individual commercial hedging
pressure (CHP) as well as aggregate CHP (\textbf{CHP}), our market wide
measure of CHP, in separate models and find that the explanatory power
of CHP at the individual commodity level decreases over the
financialisation period with the US cross-sectional average \(R^{2}\)
falling down to 20.8\% from 25.8\% in the previous period while that of
\textbf{CHP} increases (7.1\% vs 5.4\%). Financialisation therefore
seems to have had the effect of loosening the extent to which futures
risk premiums relate to commodity specific commercial hedging motives
while strengthening that to which they relate to broader dynamics such
as cross-sectional liquidity provision. The pattern is strongest for the
metals sector where the average \(R^{2}\) drops down to 19.1\% from
28.1\% between the two periods at the individual level while it rises
from 8.4\% to 15.7\% at the aggregate level. The symmetry breaks in the
crisis period, with the US cross-sectional average \(R^{2}\) at the
individual level bouncing back up (22.5\% vs 20.8\%) while that at the
aggregate level rises further (14.7\% vs 7.1\%), before eventually
resuming post-crisis with a further rise at the individual level (27.2\%
vs 22.5\%) and a drop at the aggregate level (7.5\% vs 14.7\%)
suggesting a return to legacy commodity specific dynamics.

@@

Financialisation was a global phenomenon, through the channel of index
investment it simultaneously influenced the entire spectrum of liquid
commodities; this sweeping impact spurred the transformation of
commodity markets from a collection of idiosyncratic markets into a
unified asset class with interconnectedness emerging as a crucial aspect
that warrants thorough examination.\\
We strive to address this issue by first examining the average of weekly
returns pairwise correlations between commodities across the entire
cross-section of the commodity assets considered in the study as well as
by country, sector and sub-sector. Table~\ref{tbl-correlations-periods}
shows the results for the four periods of interest with the results
reported for the entire period as well as for periods of
\textbf{backwardation} and \textbf{contango}. For the entire sample the
average pairwise correlation jumps from 0.07 in the first period to 0.16
over financialisation. This finding indicate higher interdependence
within the full cross-section of commodities starting with the
financialisation period, consistent with the empirical findings in
\citet{fryMcKibbin_evolution_2023} and
\citet{mayer_financialization_2017} as well as with the theoretical
model of \citet{basak_model_2016}. The sector analysis sheds light on
the driving force behind these changes. The US metals see the sharpest
increase in pairwise correlations going from 0.22 over the past period
to 0.56 over financialisation while the three other sectors see a much
smaller percentage increase during financialisation relative to the past
period. These results thus suggest that the US metals sector was the
driving force behind increased average pairwise correlations during
financialisation, a new observation to the best of our knowledge. The
pattern of correlations is very similar across phases of
\textbf{backwardation} and \textbf{contango} for all the sectors and
sub-sectors over the past and financialisation periods which indicates
that increased commodity interdependence over financialisation was not a
phase driven phenomenon in the same way that individual commodity
returns are, as demonstrated in Table~\ref{tbl-stats-combined}.\\
We further examine these correlations by year in
Table~\ref{tbl-correlations-years} to uncover any distinct breakpoints.
For the full sample of assets, the average pairwise correlations
increase from 0.08 to 0.12 from 2003 to 2004 seemingly driven by US
metals average pairwise correlations, more than doubling from 0.22 to
0.49, suggesting 2004 as the onset of financialization as noted in
several studies \citep{baker_financialization_2021, tang_index_2012} and
our contribution is to observe that the US metals sub-sector seems to
have played a major role in this.

@@

portfolio enables a refined examination of cross correlations across

@@

across the entire cross-section of commodity assets considered in the
study, including the GB traded base metals, increases to around 20\%
during financialisation compared to around 10\% over the previous period
with the rise driven by the metals sector at large. The sector average
for the US traded assets jumps to 31\% during financialisation from
7.5\% over the past period while for the GB traded assets the
corresponding figure increases from less than 2\%, or virtually no
explanatory power for the (US based) market factor, to almost 12\% over
the financialisation period. The results for our other risk factor
mimicking portfolios show a similar pattern and provide a deeper
insights into co-movement dynamics in this context. All three
factors---CHP, open interest and term structure---performed poorly in
the first period, indicating that the commodities with extreme factor
attributes (high or low CHP for example) were uncorrelated with the rest
of the commodity cross-section, and the overall performance consistent
across periods of \textbf{contango} and \textbf{backwardation}. The
onset of financialisation leads to a substantial improvement in the
performance of the CHP factor with its average \(R^{2}\), albeit still
small at 5.4\%, almost tripling from the past to the financialisation
period in contrast to the other factors that show weaker dynamics with a
slight decrease and a much smaller increase for the open interest and
term structure factors respectively. Here again the increase in pricing
performance for the CHP factor is driven by the metals sector,
particularly the precious sub-sector with the average \(R^{2}\) for the
group jumping to almost 26\% during financialisation from 1.2\% over the
past period. The corresponding figures of 22\% and 1.1\% for the entire
US metals sector stand in sharp contrast with the agricultural sector,
which exhibits no notable change, and the energy sector, where pricing
performance drops substantially from 9.2\% down to 1.6\%. The
improvement in the CHP factor's pricing performance remarkably comes
predominantly from the long leg (low CHP assets: Keynesian ``normal
backwardation'') of the factor\^{}\{Results available from the authors
upon request.\} with the corresponding mimicking portfolio's performance
itself driven by the metals, the precious sub-sector in particular. The
average \(R^{2}\) for the group jumps to 55\% during financialisation
from less than 12\% over the past period while the short leg sees a far
smaller increase, up to 8.9\% from 3.3\% over the past period.
Interestingly, although the term structure factor itself only sees a
very modest increase, its short leg (low roll-yield assets: Working's
backwardation) shows a sizeable rise in pricing performance, yet again
driven the metals sector, with the average \(R^{2}\) for the group's US
traded assets jumping to 28.2\% during financialisation from 6.2\% over
the past period. The pattern of results for both factors extends to the
GB trade assets and suggest a convergence of Keneysian and Working's
term structure perspective over the financialisation period.\\

>>>>

\begin{abstract}
In the early 2000s institutional investors entered the commodity futures
markets en masse with passive, long only, index type positions in sharp
contrast with those typically assumed by traditional expert
participants. A heated public debate soon erupted over the perceived
consequences of the phenomenon---commonly referred to as
``financialisation''---and, in response to immediate policy concerns,
the matter was thrust onto the academic sphere as a burning issue. With
the benefit of hindsight it now seems that the academic debate was
framed rather narrowly, used contentious research methods, and
eventually led to regulatory changes that were therefore perhaps
unwarranted. In contrast, we take a broader approach where we consider a
large cross-section of liquid commodities as suggested by the nature of
financialisation that comprehends commodity futures as an asset class.
We examine the cross-sectional interconnectedness of these assets and
uses a bespoke asset pricing factors based framework to study the issue
through the lens of co-movement. We find that the phenomenon had
ontological consequences for the commodity complex with its impact
extending beyond the mechanical effects induced by indexation. The onset
of the financial crisis and the monetary policy regimes that followed,
on the other hand, seem to have set off a motion of reversion to legacy
pre-financialisation fundamentals.
\end{abstract}

@@

The last twenty years have seen major upheavals in the commodity markets
with the onset of financialisation\footnote{Starting around 2004, with a
  view of commodity markets as an emerging asset class and in an effort
  to hedge against inflation and seek diversification benefits
  \citep{buyuksahin_speculators_2014, singleton_investor_2013},
  institutional investors sent forth an unprecedented flow of capital
  into commodity futures markets. This phenomenon is commonly referred
  to as the financialisation of commodities
  \citep{domanski_financial_2007}.} as well as the period of the 2008
financial crisis and its aftermath. Financialisation appears to have
affected multiple economic sectors, including agriculture and energy,
and its impact has been hotly debated in the policy, legislative,
regulatory, and academic spheres. A rigorous legislative scrutiny
accompanied by a vigorous policy and academic debate across a number of
disciplines about the perceived effects on commodity prices of the
unprecedented inflow of institutional funds brought into commodity
futures markets by new index type investors\footnote{See section
  Section~\ref{sec-background} for a detailed discussion.} eventually
led to regulatory changes in several individual markets.\\
In response to immediate policy concerns, the academic debate was
initially framed rather narrowly around adequacy of speculation---the
burning issue of the day---at the individual commodity level and the
ensuing academic analysis focused on the more mechanical effects of
financialisation. With the benefit of hindsight, this approach seems to
lack a fundamental understanding of the phenomenon it endeavoured to
study and which consequences it attempted to address.\\
Financialisation was a phenomenon global in nature that, through the
channel of index investment, affected the whole cross-section of liquid
commodities at once, eventually spurring its transition away from a
collection of idiosyncratic markets to its emergence as an asset class.
Focusing on individual markets in this context thus seems limited. Most
of the studies underpinning the regulatory response to financialisation
took the form of pairwise correlation, causation and to some extent
co-integration analysis, all of which are idiosyncratic in nature for
they study single pairs of individual assets. Besides, some of these
techniques since proved controversial and seem to produce results the
quality of which may fall short of the standards that may reasonably be
expected to lay the ground for legislative action. Although appealing,
for it comes with the promise of prompt ``significant'' results, this
approach misses a critical corollary of the substantial advent of a new
category of players in a (set of) market(s): the alteration of
fundamentals it potentially induces, best observed in this context
through the lens of commercial hedging pressure
\citep{keynes_treatise_1930, hicks_value_1939}.\\
While essential to the understanding of the phenomenon, the global
nature of financialisation and its impact on market fundamentals
therefore appear to have been originally overlooked and another
approach, perhaps more refined, seems warranted.

@@

These entered the commodity spectrum en masse with across-the-board
index type positions propping the transition of the complex from a
collection of idiosyncratic markets towards a more homogeneous ensemble
akin to an asset class. In this context of heightened cross-asset market
participation the issue of interconnectedness stands out as paramount
and we strive to investigate it by first examining the correlation
structure of our sample set of assets at the global, country, sector and
subsector levels. Our results indicate higher interdependence within the
full cross-section of commodities over the financialisation period,
consistent with both the empirical and theoretical literature. The
increase appears to be driven by the US metals sector, a new observation
to the best of our knowledge, and, unlike individual commodity returns,
does not seem to be regime-driven, as a similar pattern of results is
observed across phases of \textbf{backwardation} and \textbf{contango}.

@@

The returns pull-back continues into the post-crisis period,
interestingly amidst a lower volatility environment, suggesting that
investment outflows from the commodity futures complex have had a lower
impact on volatility than the corresponding inflows witnessed during
financialisation. The pattern of reversion toward the
pre-financialisation Keynesian paradigm continues stronger with mean
returns higher during phases of \textbf{backwardation} relative to
\textbf{contango} and the difference more pronounced than over the
crisis period. In contrast to the crisis period when they seemed to be
converging, the pricing performance of CHP and \textbf{CHP} now part
way, seemingly reverting to the Keynesian paradigm observed
pre-financialisation, further suggesting a reversion to legacy commodity
specific dynamics. The pull back extends to the correlation structure
with an across-the-board decrease in average pairwise correlations down
to levels below those observed during financialisation. The average
market factor beta for the entire cross-section shows a similar pattern,
dropping down below the level observed during financialization with the
decline seemingly driven by the metals sector, particularly the GB
traded segment. The pattern extends to the pricing performance of all
factors that drops below levels observed during financialisation.\\
Moreover, both the average correlations and the pricing performance of
the market factor are considerably stronger during \textbf{contango}
phases, unlike previous periods, which showed no substantial difference
between regimes with the pattern particularly pronounced for metal
assets globally, both US and GB traded. These results document the
emerging role/dominance of the \textbf{contango} regime post-crisis and
could provide a valuable basis for further research.

@@

A world-wide debate ensued about the role of index funds in commodity
markets. The first responses to the 2007/2008 crisis of escalating food
and energy prices took the form of policy reports, many of which
reasoned that the growth of commodity index funds came along with an
influx of largely speculative capital that was responsible for driving
commodity prices beyond their historic highs
\citep{deschutter_food_2010, gilbert_speculative_2010, herman_not_2011, schumann_hunger_2011, unctad_global_2009}.\\
Early US Senate investigations on the matter drew on pricing and trading
data supplied by the CFTC as well as interviews with numerous experts
including hedge fund manager Michael Masters who linked the growing
presence of index investors in US agricultural and energy markets to the
observed surges in both futures and cash prices
\citep{masters_testimony_2008, masters_accidental_2008}. This contention
has since come to be known as the ``Masters' hypothesis'' and was
largely endorsed by the final US Senate report on the issue
\citep{senate_excessive_2009}.\\
At this stage, the question thrust on the academic sphere was whether
``excessive speculation'' (understood as the market activity peculiar to
index type investors) was linked to escalating energy and agricultural
commodity prices. A number of ensuing academic studies, most of which
used commodity specific correlation and causality based analysis have
since disproved this contention: \citet{irwin_devil_2009},
\citet{sanders_adequacy_2010}, \citet{sanders_impact_2011},
\citet{sanders_new_2011}, \citet{irwin_commodity_2013},
\citet{brunetti_commodity_2014}, \citet{hamilton_effects_2015} and
\citet{bruno_financialisation_2017} investigated the issue in the
context of the agricultural markets; meanwhile,
\citet{buyuksahin_speculators_2011}, \citet{tokic_speculation_2012},
\citet{fattouh_role_2013}, \citet{kilian_role_2014},
\citet{knittel_simple_2016} and \citet{manera_modelling_2016} examined
the energy markets while \citet{bohl_does_2013}, \citet{kim_does_2015}
and \citet{boyd_prevalence_2016} studied both energy and agricultural
markets and \citet{irwin_index_2011},
\citet{irwin_financialisation_2012} as well as
\citet{stoll_commodity_2011} examined the commodity markets in general.
Other studies have underlined various alterations in commodity pricing
dynamics over the period. \citet{gilbert_role_2014} reject the view that
financialisation has not had any effects in the grains markets and
demonstrate that trades originated by financial market participants, and
specifically index investors, can move prices but tend to be typically
volatility-reducing. \citet{juvenal_speculation_2015} contend that the
oil price increase between 2004 and 2008 was mainly driven by the
strength of global demand but that the financialisation process of
commodity markets also played a role. Likewise,
\citet{henderson_new_2015} show that non-information-based financial
investments have important impacts on commodity prices.
\citet{cheng_financialisation_2014}, on the other hand, show that in the
case of index commodities, financialisation has transformed the risk
sharing, and information discovery functions of commodity futures
markets.

@@

For each period, the return series for the individual US-traded assets
are regressed on both the relative change on their respective CHP series
and that on the cross-sectional CHP independently. The resulting
\(R^{2}\) values are then averaged by sector and subsector.
Additionally, pairwise correlations are computed for each combination of
assets across the periods of interest and for each individual year
within the full sample period. These correlation coefficients are then
averaged across the entire cross-section of assets, as well as by
country, sector, and subsector. For each period, the coefficients are
also calculated and averaged across phases of low and high
cross-sectional CHP.

@@

(XLME)---show a lower figure with five og them significant---Crude

@@

The onset of financialisation thus seems to have engendered a change in
the nature of hedger's behaviour, at the aggregate level in particular,
taking it away from the traditional Keynesian view of risk shifting.
This phenomenon has been noted in earlier theoretical and empirical work
\citep{danthine_information_1978, stout_why_1998} in different contexts
from which two competing models of behaviour have emerged.\\
The information arbitrage model
\citep{grossman_impossibility_1980, danthine_information_1978, kyle_informed_1989}
implies that hedgers may often be seeking out counterparties to trade
with rather than being purely passive. This issue is also raised in both
\citet{cheng_financialisation_2014} and \citet{stulz_rethink_1996} who
point out that hedgers may be taking a view on prices just as
speculators do. In fact, by hedging away some of their risk, hedgers are
able to speculate more heavily based on their disagreement against
speculators regarding futures price movement
\citep{simsek_speculation_2013}, which fits into a heterogeneous
expectation theory of speculation\footnote{\citet{hirshleifer_speculation_1975};
  \citet{hirshleifer_reply_1976}; \citet{hirshleifer_theory_1977}}. As
\citet{stout_why_1998} points out, this disagreement-based theory of
trading is one of the main reason the public and the law disapproves
speculation as this form of trading is regarded as
non-productive\footnote{\citet{duffie_challenges_2014} also discusses
  some of the challenges faced by a policy treatment of speculative
  trading motivated by differences in beliefs.}.\\
In the second model hedgers are seen as liquidity providers as in
\citet{kang_tale_2020} with this explanation also consistent with the
nature of financialisation which led to the arrival of long term, long
only investors who require substantial liquidity to roll-over their
positions. Table~\ref{tbl-time-reg-US-CHP} supports this view. We
regress commodity futures front returns on individual commercial hedging
pressure (CHP) as well as aggregate CHP (\textbf{CHP}), our market wide
measure of CHP, in separate models and find that the explanatory power
of CHP at the individual commodity level decreases over the
financialisation period with the US cross-sectional average \(R^{2}\)
falling down to 20.8\% from 25.8\% in the previous period while that of
\textbf{CHP} increases (7.1\% vs 5.4\%). Financialisation therefore
seems to have had the effect of loosening the extent to which futures
risk premiums relate to commodity specific commercial hedging motives
while strengthening that to which they relate to broader dynamics such
as cross-sectional liquidity provision. The pattern is strongest for the
metals sector where the average \(R^{2}\) drops down to 19.1\% from
28.1\% between the two periods at the individual level while it rises
from 8.4\% to 15.7\% at the aggregate level.

@@

Financialisation was a global phenomenon, through the channel of index
investment it simultaneously influenced the entire spectrum of liquid
commodities; this sweeping impact spurred the transformation of
commodity markets from a collection of idiosyncratic markets into a more
homogeneous collection akin to an asset class. In this context
interconnectedness emerges as a crucial aspect that warrants thorough
examination.\\
We strive to address this issue by first examining the average of weekly
returns pairwise correlations between commodities across the entire
cross-section of the commodity assets considered in the study as well as
by country, sector and sub-sector. Table~\ref{tbl-correlations-periods}
shows the results for the four periods of interest with the results
reported for the entire period as well as for periods of
\textbf{backwardation} and \textbf{contango}. For the entire sample the
average pairwise correlation jumps from 0.07 in the first period to 0.16
over financialisation. This finding indicates higher interdependence
within the full cross-section of commodities starting with the
financialisation period, consistent with the empirical findings in
\citet{fryMcKibbin_evolution_2023} and
\citet{mayer_financialization_2017} as well as with the theoretical
model of \citet{basak_model_2016}. The sector analysis sheds light on
the driving force behind these changes. The US metals see the sharpest
increase in pairwise correlations going from 0.22 over the past period
to 0.56 over financialisation while the three other sectors see a much
smaller percentage increase during financialisation relative to the past
period. These results thus suggest that the US metals sector was the
driving force behind increased average pairwise correlations during
financialisation, a new observation to the best of our knowledge. The
pattern of correlations is very similar across phases of
\textbf{backwardation} and \textbf{contango} for all the sectors and
sub-sectors over the past and financialisation periods which indicates
that increased commodity interdependence over financialisation was not a
phase driven phenomenon in the same way that individual commodity
returns are, as demonstrated in Table~\ref{tbl-stats-combined}.\\
We further examine these correlations by year in
Table~\ref{tbl-correlations-years}. For the full sample of assets, the
average pairwise correlations remain in the 0.06--0.08 range from 1998
through 2003, before suddenly increasing to 0.12 from 2003 to 2004,
suggesting 2004 as a distinct break point corresponding to the onset of
financialisation, as noted in several studies
\citep{baker_financialization_2021, tang_index_2012}. We further observe
that the US metals appear to have played a major role in this as the
increase seems driven by the sector which average pairwise correlation
more than doubles (0.49 vs 0.22) over the period.

@@

portfolio enables a refined examination of cross-correlations across

@@

across the entire cross-section of commodity assets, including the GB
traded base metals, increases to around 20\% during financialisation
compared to around 10\% over the previous period with the rise driven by
the metals sector at large. The sector average for the US traded assets
jumps to 31\% during financialisation from 7.5\% over the past period
while for the GB traded assets the corresponding figure increases from
less than 2\%, or virtually no explanatory power for the (US based)
market factor, to almost 12\% over the financialisation period. The
results for our other risk factor mimicking portfolios show a similar
pattern and provide a deeper insights into co-movement dynamics in this
context. All three factors---CHP, open interest and term
structure---performed poorly in the first period, indicating that the
commodities with extreme factor attributes (high or low CHP for example)
were uncorrelated with the rest of the commodity cross-section, and the
overall performance consistent across periods of \textbf{contango} and
\textbf{backwardation}. The onset of financialisation leads to a
substantial improvement in the performance of the CHP factor with its
average \(R^{2}\), albeit still small at 5.4\%, almost tripling from the
past to the financialisation period in contrast to the other factors
that show weaker dynamics with a slight decrease and a much smaller
increase for the open interest and term structure factors respectively.
Here again the increase in pricing performance for the CHP factor is
driven by the metals sector, particularly the precious sub-sector with
the average \(R^{2}\) for the group jumping to almost 26\% during
financialisation from 1.2\% over the past period. The corresponding
figures of 22\% and 1.1\% for the entire US metals sector stand in sharp
contrast with the agricultural sector, which exhibits no notable change,
and the energy sector, where pricing performance drops substantially
from 9.2\% down to 1.6\%. The improvement in the CHP factor's pricing
performance remarkably comes predominantly from the long leg (low CHP
assets: Keynesian ``normal backwardation'') of the factor\footnote{Results
  available from the authors upon request.} with the corresponding
mimicking portfolio's performance itself driven by the metals, the
precious sub-sector in particular. The average \(R^{2}\) for the group
jumps to 55\% during financialisation from less than 12\% over the past
period while the short leg sees a far smaller increase, up to 8.9\% from
3.3\% over the past period. Interestingly, although the term structure
factor itself only sees a very modest increase, its short leg (low
roll-yield assets: Working's backwardation) shows a sizeable rise in
pricing performance, yet again driven the metals sector, with the
average \(R^{2}\) for the group's US traded assets jumping to 28.2\%
during financialisation from 6.2\% over the past period. The pattern of
results for both factors extends to the GB traded assets and suggest a
convergence of Keneysian and Working's term structure perspective over
the financialisation period.\\
